Generative Simplex Mapping: Nonlinear Endmember Extraction and Unmixing for Hyperspectral Imagery

### 2024-08-01 paper/paper.tex
more updates to the paper

%=================================================================
% Add packages and commands here. The following packages are loaded in our class file: fontenc, inputenc, calc, indentfirst, fancyhdr, graphicx, epstopdf, lastpage, ifthen, float, amsmath, amssymb, lineno, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, tabto, xcolor, colortbl, soul, multirow, microtype, tikz, totcount, changepage, attrib, upgreek, array, tabularx, pbox, ragged2e, tocloft, marginnote, marginfix, enotez, amsthm, natbib, hyperref, cleveref, scrextend, url, geometry, newfloat, caption, draftwatermark, seqsplit
% cleveref: load \crefname definitions after \begin{document}

@@

1. Overview of problem

Hyperspectral imaging has emerged as a keystone technology in remote sensing where the ability to discern slight variations in high resolution spectra facilitates a plethora of critical applications from land use assessment... etc. 
- environmental monitoring, resource management, risk assessment, biodiversity conservation, sustainable agriculture,
- Many hyperspectral remote sensing missions have already been launched with many more in development
    - EnMAP (launched 2022) is german satellite with remote sensing imager
    - PRISMA (launched 2019) is italian remote sensing satellite
    - HISUI (launched 2019) is japanese HSI on international space station
    - CHIME (future launch) - Copernicus Hyperspectral Imaging mission for the environment from the european space agency
    - SBG (future launch) - surface biology and geology - NASA mission
    - SHALOM (future launch) - Joint israeli-italian mission
    - HyspIRI (future launch) - italian missions
- Additionally, continued improvement to HSI technologies have dramatically reduced their size enabling their inclusion to the payloads of small unmanned aerial vehicles (UAV).

- Key challenge: how to make sense of big data
- Common approaches involve computation of spectral indices like the NDVI
    - These are nice but can be challenging to compare between instruments due to varibility in spectral bands and sensor response
    - They also throw away
- A second common approach is to directly map spectral data to parameters of interest using in situ reference data 
    - This is popular for water quality assessment where parameters like CDOM, etc. track water quality
    - The key limitation is the availability of coincident reference data. This can requires decades of aligned observations to provide sufficient data volumes for statistical/machine learning models
- In light of these difficulties, unsupervised methods which extract relevant information from HSI themselves can be advantageous
- Reality: pixels captured by remote sensing platforms have spatial extents which typically cover multiple unique sources. 
- overview of spectral unmixing
- relevance to source identification in remote sensing, contaminant identication 
- relevance to source apportionment in air quality monitoring

@@

\begin{figure}[H]
\includegraphics[width=\columnwidth]{paper/figures/methods/gsm/gsm-diagram.pdf}
\caption{Illustration of the GSM. Points in the latent space (green dots) are uniformly distributed at the nodes of a regular grid on the n-simplex whose barycentric coordinates are to be interpreted as the relative abundance of n-many unique sources $z_1, ... z_{N_v}$. Here, $N_v=3$ has been chosen for illustrative purposes. These nodes are mapped into (spectral) data space via the mapping $\psi(z)$ utilizing $M$-many radially symmetric basis functions shown in red. The trained GSM also enables modelling of spectral variability via the precision parameter $\beta$ shown here as a light blue band around the estimated spectra in the data space.\label{fig:gsm-diagram}}
\end{figure}

@@

\begin{figure}[H]
\includegraphics[width=\columnwidth]{paper/figures/methods/usgs/usgs-dataset.pdf}
\caption{Synthetic dataset from USGS spectra. \label{fig:usgs-data}}
\end{figure}

@@

\begin{figure}[H]
\includegraphics[width=\columnwidth]{paper/figures/methods/robot-team/robot-team-overview.pdf}
\caption{Robot team dataset. \label{fig:robotteam-data}}
\end{figure}

@@

\begin{figure}[H]
\includegraphics[width=\columnwidth]{paper/figures/results/usgs/fit-comparison.pdf}
\caption{Comparison of GSM against NMF on simulated USGS dataset. \label{fig:usgs-fits}}
\end{figure}

\begin{figure}[H]
\includegraphics[width=\columnwidth]{paper/figures/results/usgs/extracted-endmembers.png}
\caption{Endmembers extracted using GSM for simulated USGS dataset with SNR$=20$. \label{fig:usgs-endmembers}}
\end{figure}

@@

Exploratory data analysis: 
\begin{figure}[H]
\includegraphics[width=0.60\columnwidth]{paper/figures/results/robot-team/pca-variance.png}
\caption{Explained variance of PCA components for robot team dataset. A red horizontal line is superimposed on the plot indicating an explained variance of $1\%$. \label{fig:robot-team-pca}}
\end{figure}

@@

\begin{figure}[H]
\begin{adjustwidth}{-\extralength}{0cm}
\centering
\includegraphics[width=1.25\columnwidth]{paper/figures/results/robot-team/extracted-endmembers.pdf}
\end{adjustwidth}
\caption{Nonlinear GSM applied to water spectra: \textbf{(a)} Spectra corresponding to maximum abundances for each vertex in the trained GSM identified for as water, near-shore vegetation, and rhodamine dye sources. \textbf{(b)} The original hyperspectral datacube segmented according to the relative abundance of each endmember. Each water pixel defined by an NDWI $\geq 0.25$ is colored by smoothing interpolating between red, green, and blue colors corresponding to the relative abundance estimated for rhodamine, vegetation, and water spectra.}
\label{fig:robot-team-endmembers}
\end{figure}

\newpage
\begin{figure}[H]
\begin{adjustwidth}{-\extralength}{0cm}
\centering
\includegraphics[width=\columnwidth]{paper/figures/results/robot-team/endmember-abundances.pdf}
\caption{Endmember distributions: \textbf{(a)} The distribution of abundance for the water class. This source dominates in the center of the water and decreases towards the edge of the pond where surface vegetation begins to dominate the reflectance signal. We note that the water abundance decreases near  the edge of the rhodamine plume reflecting dye mixing and diffusion. \textbf{(b)} The distribution of vegetation. This signal includes filamentous blue-green algae observed to accumulate in shallow waters near the shore. \textbf{(c)} The rhodamine dye plume extent segmented from the HSI. The total area for near-shore vegetation and rhodamine are estimated to be $378.6$ $\text{m}^2$ and $255.7$ $\text{m}^2$ respectively.}
\end{adjustwidth}
\label{fig:endmember-abundance-dist}
\end{figure}  
\newpage

\begin{figure}[H]
\begin{adjustwidth}{-\extralength}{0cm}
\centering
\includegraphics[width=1.25\columnwidth]{paper/figures/results/robot-team/plume-evo.pdf}
\end{adjustwidth}
\caption{Rhodamine plume evolution: Using the trained GSM we can track the dispersion of the rhodamine dye plume between successive drone flights. \textbf{(a)} The initial plume distribution after release. Here the dye subsumes an area of $255.7$ $\text{m}^2$. \textbf{(b)} The same plume imaged 15 minutes later now extends across an area of $571.8$ $\text{m}^2$}
\label{fig:plume-evo}
\end{figure}

@@

\noindent 
\begin{tabular}{@{}ll}
MDPI & Multidisciplinary Digital Publishing Institute\\
DOAJ & Directory of open access journals\\
TLA & Three letter acronym\\
LD & Linear dichroism
\end{tabular}
}

@@

\bibliography{paper/references}
\PublishersNote{}

>>>>

%=================================================================
% Add packages and commands here. The following packages are loaded in our class file: fontenc, inputenc, calc, indentfirst, fancyhdr, graphicx, epstopdf, lastpage, ifthen, float, amsmath, amssymb, lineno, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, tabto, xcolor, colortbl, soul, multirow, microtype, tikz, totcount, changepage, attrib, upgreek, array, tabularx, pbox, ragged2e, tocloft, marginnote, marginfix, enotez, amsthm, natbib, hyperref, cleveref, scrextend, url, geometry, newfloat, caption, draftwatermark, seqsplit
% cleveref: load \crefname definitions after \begin{document}
\graphicspath{{figures/}}

@@

Hyperspectral imaging has emerged as a keystone technology in remote sensing where the ability to discern variations between high resolution spectra facilitates a plethora of critical applications from environmental monitoring and risk assessment to biodiversity convservation and sustainable agriculture. In recent years, many remote sensing platforms have been deployed with hyperspectral imaging payloads such as the Italian PRISMA mission launched in 2019 and the more recent German EnMAP launched in 2022 (\textcolor{red}{add citations}). Many future missions also plan to incorporate hyperal imaging capabilities such as the European Space Agency's CHIME which will include over $200$ bands spanning visible, near infrared (NIR), and short-wave infrared (SWIR) wavelengths (\textcolor{red}{add citations}). Additionally, the continued development of hyperspectral imaging technologies have resulted in a considerable reduction in size enabling their inclusion in the payloads of small unmanned aerial vehicles (UAV) (\textcolor{red}{add citations}). Despite these promising advances, the substantial increase in data volume generated by hyperspectral imagers as compared to their multi-spectral counterparts poses significant challenges to real-time analysis at scale.

Many approaches have been developed to make sense of hyperspectral images (HSI). For example, spectral indices such as the popular normalized difference vegetation index (NDVI) can be computed by taking ratios of spectral bands tailored to track specific reflectance characteristics (\textcolor{red}{add citations}). These indices have the advantage of being easy to compute but suffer from significant inter-instrument variability while ingoring most of the captured wavelength bins (\textcolor{red}{add citations}). An alternative approach is to pair HSI data with in-situ measurements to enable supervised models which map spectra directly to parameters of interest. However, this approach relies on serendipitous satellite overpasses above sensing sites to generate sufficient quantities of aligned data for model training and evaluation. For example, Aurin et al. combined data from over 30 years of oceanographic field campaigns with paired satellite imagery to develop robust models for the inversion of key water quality indicators such as colored dissolved organic matter (CDOM) \cite{aurin2018remote}. This approach can be accelerated by combining UAV-based hyperspectral imaging with rapid in-situ data collection using autonomous boats \cite{robot-team-1, robot-team-2}. Nevertheless, these supervised methods rely on \textit{a priori} knowledge of expected sources in order to select appropriate reference sensors for data collection. In light of these limitations, unsupervised methods are needed which can expediently extract source signatures from high dimensional HSI.

The spatial resolution of hyperspectral imagers generally results in mixed HSI pixels comprised of many individual sources called endmembers. The task of unsupervised source identification using HSI data therefore comprises two steps: endmember extraction, and abundance estimation. Techniques such as vertex component analysis (VCA), the pixel purity index (PPI), and N-FINDR, solve this first task by identifying endmember spectra from HSI assuming the presense of some pure (unmixed) pixels (\textcolor{red}{add citations}). By further assuming a linear mixing model (LMM) whereby observed spectra are described by a linear combination of endmembers with non-negative abundances, HSI can be unmixed to obtain abundance maps using a variety of techniques such as constrained least-squares.

Once extracted, HSI can be unmixed to estimate individual endmember abundances using a variety of techniques such as constrained least-squares optimization (\textcolor{red}{add citations}). Alternatively,

- Reality: pixels captured by remote sensing platforms have spatial extents which typically cover multiple unique sources. 
- overview of spectral unmixing
- relevance to source identification in remote sensing, contaminant identication

@@

\begin{figure}[H]
\includegraphics[width=\columnwidth]{methods/gsm/gsm-diagram.pdf}
\caption{Illustration of the GSM. Points in the latent space (green dots) are uniformly distributed at the nodes of a regular grid on the n-simplex whose barycentric coordinates are to be interpreted as the relative abundance of n-many unique sources $z_1, ... z_{N_v}$. Here, $N_v=3$ has been chosen for illustrative purposes. These nodes are mapped into (spectral) data space via the mapping $\psi(z)$ utilizing $M$-many radially symmetric basis functions shown in red. The trained GSM also enables modelling of spectral variability via the precision parameter $\beta$ shown here as a light blue band around the estimated spectra in the data space.\label{fig:gsm-diagram}}
\end{figure}

@@

\begin{figure}[H]
\includegraphics[width=\columnwidth]{methods/usgs/usgs-dataset.pdf}
\caption{Synthetic dataset from USGS spectra. \label{fig:usgs-data}}
\end{figure}

@@

\begin{figure}[H]
\includegraphics[width=\columnwidth]{methods/robot-team/robot-team-overview.pdf}
\caption{Robot team dataset. \label{fig:robotteam-data}}
\end{figure}

@@

\begin{figure}[H]
\includegraphics[width=\columnwidth]{results/usgs/fit-comparison.pdf}
\caption{Comparison of GSM against NMF on simulated USGS dataset. \label{fig:usgs-fits}}
\end{figure}

\begin{figure}[H]
\includegraphics[width=\columnwidth]{results/usgs/extracted-endmembers.png}
\caption{Endmembers extracted using GSM for simulated USGS dataset with SNR$=20$. \label{fig:usgs-endmembers}}
\end{figure}

@@

Exploratory data analysis: 
\begin{figure}[H]
\includegraphics[width=0.60\columnwidth]{results/robot-team/pca-variance.png}
\caption{Explained variance of PCA components for robot team dataset. A red horizontal line is superimposed on the plot indicating an explained variance of $1\%$. \label{fig:robot-team-pca}}
\end{figure}

@@

\begin{figure}[H]
\begin{adjustwidth}{-\extralength}{0cm}
\centering
\includegraphics[width=1.25\columnwidth]{results/robot-team/extracted-endmembers.pdf}
\end{adjustwidth}
\caption{Nonlinear GSM applied to water spectra: \textbf{(a)} Spectra corresponding to maximum abundances for each vertex in the trained GSM identified for as water, near-shore vegetation, and rhodamine dye sources. \textbf{(b)} The original hyperspectral datacube segmented according to the relative abundance of each endmember. Each water pixel defined by an NDWI $\geq 0.25$ is colored by smoothing interpolating between red, green, and blue colors corresponding to the relative abundance estimated for rhodamine, vegetation, and water spectra.}
\label{fig:robot-team-endmembers}
\end{figure}

\newpage
\begin{figure}[H]
\begin{adjustwidth}{-\extralength}{0cm}
\centering
\includegraphics[width=\columnwidth]{results/robot-team/endmember-abundances.pdf}
\caption{Endmember distributions: \textbf{(a)} The distribution of abundance for the water class. This source dominates in the center of the water and decreases towards the edge of the pond where surface vegetation begins to dominate the reflectance signal. We note that the water abundance decreases near  the edge of the rhodamine plume reflecting dye mixing and diffusion. \textbf{(b)} The distribution of vegetation. This signal includes filamentous blue-green algae observed to accumulate in shallow waters near the shore. \textbf{(c)} The rhodamine dye plume extent segmented from the HSI. The total area for near-shore vegetation and rhodamine are estimated to be $378.6$ $\text{m}^2$ and $255.7$ $\text{m}^2$ respectively.}
\end{adjustwidth}
\label{fig:endmember-abundance-dist}
\end{figure}  
\newpage

\begin{figure}[H]
\begin{adjustwidth}{-\extralength}{0cm}
\centering
\includegraphics[width=1.25\columnwidth]{results/robot-team/plume-evo.pdf}
\end{adjustwidth}
\caption{Rhodamine plume evolution: Using the trained GSM we can track the dispersion of the rhodamine dye plume between successive drone flights. \textbf{(a)} The initial plume distribution after release. Here the dye subsumes an area of $255.7$ $\text{m}^2$. \textbf{(b)} The same plume imaged 15 minutes later now extends across an area of $571.8$ $\text{m}^2$}
\label{fig:plume-evo}
\end{figure}

@@

\noindent
\begin{tabular}{@{}ll}
PRISMA & Hyperspectral Precuror of the Application Mission \\
EnMAP & Environmental Mapping and Analysis Program \\
CHIME & Copernicus Hyperspectral Imaging Mission for the Environment \\
NIR & Near Infrared \\
SWIR & Short-wave Infrared \\
UAV & Unmanned Aerial Vehicle \\
HSI & Hyperspectral Image \\
NDVI & Normalized Difference Vegetation Index \\
VCA & Vertex Component Analysis\\
PPI & Pixel Purity Index
\end{tabular}
}

@@

\bibliography{./references.bib}
\PublishersNote{}
